---
title: "Check-in 1"
author: "Brendan Dishion"
date: "2024-01-29"
output: html_document
ormat: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

1.  This is a code chunk that opens the `mtcars` dataset, which contains information about on 32 cars from a 1974 issue of the magazine *Motor Trend* and creates a scatter plot with horsepower `hp` on the x-axis and miles-per-gallon `mpg` on the y-axis. Try running this code and Rendering/Knitting your Rmd document. If you have done this correctly, RStudio should produce an html document.

```{r, eval = T}
data(mtcars)
plot(mtcars$hp, mtcars$mpg)
```

2.  Using this scatter plot, describe the relationship between `mpg` and `hp`.
 
    After reviewing the scatter plot, from the data we can conclude that
    cars with fewer horsepower tend to get better mpg than cars with 
    a higher horsepower. This makes sense since vehicles with more horsepower
    require mores gas for the engine to run optimally. From the data
    we can safely conclude that cars between 50-100 hp get the best mileage while
    cars with 200-300hp generally get the worst gas mileage, but will be much faster.


3.  Create your own code chunk below this question containing the line of code `print("Hello MATH167R!")`. Then Render/Knit your Rmd document. If you have done this correctly, RStudio should produce an html document.

```{r, eval = T}
print("Hello MATH167R!")
```

