---
title: "Check-in 6"
author: "Brendan Dishion"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, **follow the instructions below and use R Markdown to create a pdf document with your code and answers to the following questions on Gradescope.** You may find a template file by clicking "Code" in the top right corner of this page.

1. The following code downloads two sequences of 256 digits. One of them was generated by asking ChatGPT to generate to "sample 1000 random digits from 0 to 9 with replacement" (I only got 256). The other was generated using the R command `sample(0:9, 256, replace = T)`. Which is from R and which is from ChatGPT? Explain and justify your answer.

```{r, warning = F, message = F}
A <- read.csv("https://math167r-s24.github.io/static/digits-a.csv")
B <- read.csv("https://math167r-s24.github.io/static/digits-b.csv")
```

```{r}
sample_return <- function(sample_size) {
  return(summary(sample(0:9, sample_size, replace = TRUE)))
}

summary(A)
cat("\n")
summary(B)
cat("\n")
mean(replicate(1000, sample_return(256)))
```

Based on the code above, I create a function that returns the summary of a sample from 0-9 with replacement. When taking a look at the summary for A, it's clear that the mean is approximately 4.5. For B, the mean is approximately 4.9. When replicating the function we created, it's clear to see that R will produce a mean close to 4.5. So A is produced from R, and B is from ChatGPT. The replication repeats 1000 times, so this is accurate.

Justification: Sequence A most likely originated from R due to its statistics aligning closely with the expected mean of 4.5 for a uniform distribution of digits from 0 to 9. This is evident from the mean value being close to 4.5, which reflects consistent behavior from R's random number generation process.

Sequence B is more likely to have been generated by ChatGPT. The slightly higher mean of approximately 4.9 suggests a deviation from the expected mean.The means of the replicated samples tend to converge around 4.5, consistent with R's uniform distribution, while Sequence B maintains a slightly higher mean value. 

Although said, R can deviate from this value and can print values like 4.8, but for sake of assumption it's fair to say that the mean value in R will converge towards 4.5 and that sequence B most likely was generated from ChatGPT. Even though either tool could've printed these values, the expected value of the mean should be around 4.5 for samples from 0-9 with replacement in R. Therefore, sequence A is most likely generated from R and sequence B is most likely generated from ChatGPT.

