---
title: "Check-in 7"
author: "Brendan Dishion"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---


Load the flips data using the following code:

```{r, warning = F, message = F}
library(tidyverse)
flips <- read_csv("https://math167r-s24.github.io/static/flips.csv")
head(flips)
```

1. Design your own hypothesis test to try to identify the sequence of real flips. Design your own test statistic and simulate the null distribution of your test statistic. Can you identify which sequence is the real one?

```{r}
# function to count the # of alternations between 
# heads and tails in a sequence of coin flips
num_alterations <- function(sequence) {
  
  n_alterations <- 0
  
  for(i in 2:length(sequence)) {
    
    # Checking if the current element is the same as the previous one
    if(sequence[i] != sequence[i - 1]) { 
      n_alterations <- n_alterations + 1  # Increment the count of alternations
    }
  }
  return(n_alterations)
}

# Calculating the observed alternations for each sequence/column in the flips

observed_alternations <- flips |>
  
  # Applying num_alterations function to every column
  
  summarize(across(everything(), ~num_alterations(.x)))

# seed 
set.seed(1000)

flips_per_simulation <- nrow(flips)

# Simulating null distribution; generating random sequences of flips
sim_alternations <- replicate(1000, {
  
  # Randomly sample "H" or "T"
  sim_flips <- sample(c("H", "T"), flips_per_simulation, replace = TRUE)
  
  # Counting the alternations in the simulated sequence using num_alterations()
  num_alterations(sim_flips)
})

# Calculating p-values based on observed_alternations count
p_values <- sapply(observed_alternations, function(x) {
  
  # Calculates the proportion of simulations with alternations >= observed
  mean(sim_alternations >= x)
})

# Printing results
results <- bind_rows(observed_alternations, p_values = p_values)
print(results)
```

First Row: This row shows the number of alternations (changes between heads and tails) observed for each sequence.

Second Row (P-Value): The p-values indicate the probability of observing alternation counts as extreme or more extreme than the actual counts, under the assumption that the sequence was generated by a fair coin. A higher p-value suggests that the observed alternation count is typical under the null hypothesis, while a lower p-value indicates a significant deviation from the expected randomness.

Null Hypothesis (H0): The sequence displays a level of randomness consistent with what is expected from a fair coin flip, which is reflected by alternation counts that align closely with the central tendency of the simulated distribution.

Alternative Hypothesis (HA): The sequence exhibits a level of randomness that significantly deviates from that expected under the null hypothesis.

Test Statistic: The number of alternations between heads and tails within each sequence.

Calculation of P-values: The proportion of simulations where the number of alternations is as large or larger than observed provides the p-values. This comparison determines how typical the observed counts are compared to what might be expected from a fair coin.

Interpretation Based on Results: Considering the p-values, Sequence B, which is closest to a p-value of 1, is most likely to represent the real, fair sequence, indicating typical random behavior expected from a fair coin. Sequences C and D, with p-values slightly below 1, also display a high level of randomness, however slightly less typical than B.



